Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-04-17 thesis/data.tex
updated data!

Gaia provides unprecedented accuracy in proper motions and magnitudes.
As such, Gaia data is uniquely excellent to produce low-contamination
samples of likely member stars belonging to satellites. Here, we breifly
describe J+24's membership estimation and discuss how this informs our
observational knoledge of each galaxies density profile. In general,
J+24 use a Bayesian framework incorporating proper motion (PM),
colour-magnitude diagram (CMD), and spatial information to determine the
probability that a given star belongs to the satellite. J+24 extends
\citet{MV2020a} (see also \citet{pace+li2019}, \citet{battaglia+2022},
\citet{pace+erkal+li2022}, etc.).

@@

@@

Our RV methodology is

\begin{itemize}
\tightlist
\item
  Cross match all catalogues to J+24 Gaia members. If Gaia ID's are
  included, we use these. Otherwise, we match to the nearest star within
  2 arcseconds.
\item
  We combine the mean RV measurement from each study using the
  inverse-variance weighted mean and standard uncertainty.
\end{itemize}

\[
v_i = \frac{1}{\sum w}\sum_j w\ v_{i, j} \\
\delta v_i = \sqrt{\frac{1}{\sum w^2}}
\]

where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
deviation \[
s_{v}^2 = \frac{1}{\sum w} \sum w (v_{i, j} - v_i)^2
\] Note that we do not include the standard deviation if a source does
measure the velocity multiple times.

If the total standard deviation is greater than five times the velocity
standard uncertainty (reduced), then we remove the star since either the
star is a binary or contains erroneous measurements or other
uncertainties. Note that since stars typically only have very few
measurements, these choices are highly approximate and likely limited.

The combined RV likelihood is then \[
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
\] where \[
{\cal L}_{\rm RV} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
dispersion, and \(\delta v_i\) is the individual measurement
uncertainty. Typically, the velocity dispersion will dominate the
uncertainty budget here.

Finally, we need to correct the coordinate frames. We need a proper
motion estimate of each star. Since most stars have Gaia measured proper
motions, but the uncertainties on each proper motion measurement are
large, we combine the system and individual star's proper motion by
weighting the systemic proper motion by the z-score of the consistency
of the measurement with the systemic proper motion. \[
\mu_{*}' = \frac{1}{w_\star + w_{\rm sys}}\left(\mu_{\star}w_\star + \mu_{\rm sys} w_{\rm sys}\right)
\] where we model \[
{w_{\rm sys}} = P_{\rm sat}(\delta \mu ^2 + \sigma_{v, tan}^2)^{-1}
\] and the weight of the star is just the inverse variance weight plus
the gaia systematics.

The corrected radial velocities are then \[
v_z = v_{\rm los, gsr}\cos\theta  + v_{\rm tan, R} \sin\theta
\] where
\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
is the radial component of the proper motion with respect to the centre
of the galaxy. Since \(\theta \lesssim 1^\circ\), this correction is
small (\(\sim 1\)km,s\(^{-1}\)), but can prevent the introduction of a
spurious gradient on the sky \citep{WMO2008}. The uncertainty is then
the velocity uncertainty plus the distance uncertainties times the PM
uncertainty from above. We then use the \(v_z\) values for the following
modelling, however repeating with plain, solar-frame velocities does not
substantially affect the results too much .

For Sculptor, we combine radial velocity measurements

\begin{itemize}
\item
  APOGEE (crossmatched to stars meeting basic quality cuts in field)
\item
  \citet{sestito+2024a}
\item
  \citet{tolstoy+2024}
\end{itemize}

We adopt a very loose prior of \[
\mu_{v} = N(111, 10) \\
\log\sigma_{v} = N(1, 1)
\] Our resulting velocity dispersions are \$\$

\$\$ Comparing to past work, note that our values are reasonable.

\begin{itemize}
\tightlist
\item
  APOGEE
\item
  \citet{sestito+2024b}
\item
  \citet{pace+2022}
\item
  \citet{spencer+2018}
\end{itemize}

Our model here is relatively simple. Some things which we do not account
for include

\begin{itemize}
\tightlist
\item
  interstudy systematics and biases
\item
  Inappropriate uncertainty reporting.
\item
  Binarity. Binay stars are expected to induce an average velocity
  dispersion of xxx on an individual star.
\item
  Selection effects. RV studies each have their own targeted program and
\end{itemize}

We partially address several of these by also performing the analysis on
large RV samples for each galaxy. This allows for individual comparison
illustrating the differences in selection effects.

>>>>

Gaia provides unprecedented accuracy in proper motions and magnitudes.
Gaia data is uniquely excellent to produce low-contamination samples of
likely member stars belonging to satellites. Here, we breifly describe
J+24's membership estimation and discuss how this informs our
observational knoledge of each galaxies density profile. In general,
J+24 use a Bayesian framework incorporating proper motion (PM),
colour-magnitude diagram (CMD), and spatial information to determine the
probability that a given star belongs to the satellite. J+24 extends
\citet{MV2020a} (see also \citet{pace+li2019}, \citet{battaglia+2022},
\citet{pace+erkal+li2022}, etc.).

@@

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 2
arcseconds. We exclude stars not matched to Gaia for simplicity.

We combine the mean RV measurement from each study using the
inverse-variance weighted mean and standard uncertainty. \[
\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} \\
\delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
\] where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
deviation \[
s^2 = \frac{1}{\sum w_i} \sum_i w_i (v_{i} - \bar v)^2
\] We remove stars with significant velocity dispersions as measured
between or within a study: \[
s < 5\,\delta v\,\sqrt{n}
\] where \(s, \delta v, n\) are the standard deviation, standard, error,
and number of measurements for both a study (with multiple epochs) and
inter-study comparisons.

The combined RV likelihood is then \[
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
\] where \[
{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2) \\
{\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
\] where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
dispersion, and \(\delta v_i\) is the individual measurement
uncertainty. Typically, the velocity dispersion will dominate the
uncertainty budget here. We assume a halo/background velocity dispersion
of a constant \$\sigma\_\{\rm halo\} = 100 \$ km/s
\citep[e.g.][]{brown+2010}.

Finally, we need to correct the coordinate frames for the solar motion
and on-sky size of the galaxy. The first step is to subtract out the
solar motion from each radial velocity, corresponding to a typical
gradient of \textasciitilde3 km/s across the field. The next step is to
account for the slight differences in the direction of each radial
velocity. Define the \(\hat z\) direction to point parallel to the
direction from the sun to the centre of the dwarf galaxy. Then if
\(\phi\) is the angular distance between the centre of the galaxy and
the individual star, the corrected radial velocity is then \[
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
\] where
\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
is the radial component of the proper motion with respect to the centre
of the galaxy.This correction is of the order \(v_{\rm tan}\theta\) so
induces a gradient of about \(1 km/s/degree\) for sculptor
\citep[see][]{WMO2008}. The uncertainty is then the velocity uncertainty
plus the distance uncertainties times the PM uncertainty from above. We
then use the \(v_z\) values for the following modelling, however
repeating with plain, solar-frame velocities does not substantially
affect the results too much .

For the priors on the satellite velocity dispersion and systematic
velocity, we use \[
\mu_{v} = N(0, \sigma_{\rm halo}^2) \\
\sigma_{v} = U(0, 20\,{\rm km\,s^{-1}})
\] where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is the velocity
dispersion of the MW halo adopted above, a reasonable assumption for
dwarfs in orbit around the MW.

\begin{table*}[t]
\centering
\caption{Summary of velocity measurements and derived properties.}
\label{tbl:Summary-of-velocity-measurements-and-derived-properties}
\begin{tabular}{llllllll}
\toprule
 & Study & Instrument & Memb & Rep. & N$\sigma > s$ & $\bar v$ & $\sigma_v$\\
\midrule
Scl &  &  &  &  &  &  & \\
 & tolstoy+23 & VLT/FLAMES & 200 (1604) & 500 &  &  & 9.6\\
 & sestito+23a & GMOS & 2000 &  &  &  & -\\
 & walker+09 & MMFS & 2000 &  &  &  & 9.5\\
 & APOGEE & APOGEE & 200 & 200 &  & 8 & 9\\
UMi &  &  &  &  &  &  & \\
 & sestito+23b & GRACES & 6 &  &  &  & \\
 & pace+20 & Keck/DEIMOS &  &  &  & -245 & 8.6\\
 & spencer+18 & MMT/Hectoshell &  &  &  & -247 & \\
 & APOGEE & APOGEE & 100 & 100 &  &  & \\
\bottomrule
\end{tabular}
\end{table*}

For Sculptor, we combine radial velocity measurements from APOGEE,
\citet{sestito+2023a}, \citet{tolstoy+2023}, and \citet{WMO2009}.
\citet{tolstoy+2023} and \citet{WMO2009} provide the bulk of the
measurements. We find that there is no significant velocity shift in
crossmatched stars between catalogues. After crossmatching to Gaia and
excluding significant inter-study dispersions, we have a sample of XXXX
members.

We derive a systemic velocity of \(111.2\pm0.2\) km/s with velocity
dispersion \(9.67\pm0.16\) km/s. Our values are very consistent with
previous work \citep[e.g.][\citet{arroyo-polonio+2024},
\citet{tolstoy+2023}]{WMO2009}.

Finally, we attempt to fit a linear velocity gradient to Scl by adding
parameters for the gradient in \(\xi\) ande \(\eta\). We derive a
gradient of xxx and xxx (see Figure\textasciitilde FIG.), noting that
this direction is different and higher in magnitude than the proper
motion of Scl. Comparing the density difference, we find a bayes factor
of -15. Compared to past work, \citet{battaglia+2008},
\citet{arroyo-polonio+2021}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vlos_xi_eta.pdf}}
\caption[Scl velocity sample]{A plot of the corrected los velocities for
Scl binned in tangent plane coordinates. We detect a slight rotational
gradient towards the bottom right. \textbf{TODO} add a histogram of
velocities here too with fits, maybe with respect to distance.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient.pdf}}
\caption[Scl velocity gradient]{A velocity gradient in Sculptor! The
arrow marks the gradient induced by Scl's proper motion on the sky.
\textbf{TODO}: plot of velocity of Scl in direction of gradient.}
\end{figure}

\emph{Is it interesting that the velocity dispersion of Scl seems to
increase significantly with Rell?}

For UMi, we collect radial velocities from, APOGEE,
\citet{sestito+2024b}, \citet{pace+2020}, and \citet{spencer+2018}.

We shifted the velocities of \citet{spencer+2018} (\(-0.9\) km/s) and
\citet{pace+2020} (\(+1.0\) km/s) to reach the same scale based on a
crossmatch of about 200 common stars. Since the mean difference in
velocities in this crossmatch is of order 1.9 km/s, we adopt this as the
systematic LOS velocity error.

UMi interestingly has a more structured observational pattern, but does
not appear to have any significant velocity substructure.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_vlos_xi_eta.pdf}}
\caption[UMi velocity sample]{A plot of the corrected los velocities for
UMi binned in tangent plane coordinates. There is no clear rotation or
velocity gradient here. Interestingly, many velocity members are as far
as 100 arcmin away.}
\end{figure}

Our model here is relatively simple. Some things which we note as
systematics and are challenging to account fully for are

\begin{itemize}
\tightlist
\item
  Inter-study systematics and biases. While basic crossmatches and a
  simple velocity shift, combining data from multiple instruments is
  challenging.
\item
  Inappropriate uncertainty reporting. Inspection of the variances
  compared to the standard deviations within a study seems to imply that
  errors are accurately reported. APOGEE notes that their RV
  uncertainties are known to be underestimates.
\item
  Binarity. While not too large of a change for classical dwarfs, this
  could inflate velocity dispersions of about 9 km/s by about 1 km/s
  \citet{spencer+2017}. Thus, our measurement is likely slightly
  inflated given the high binarity fractions measured in classical
  dwarfs \citep[\citet{spencer+2018}]{arroyo-polonio+2023}.
\item
  Selection effects. RV studies each have their own selection effects. I
  do not know how to correct for this.
\end{itemize}

Because the derived parameters are similar for the two different larger
surveys we consider for UMi and Scl, we note that many of these effects
are likely not too significant (with the exception of the systemic
motion of UMi.)

@@

### 2025-04-17 thesis/introduction.tex
updated data!

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/power_spectrum.png}
\caption[Cosmological Power Spectrum]{figure 1 from
\citet{bechtol+2022}.}\label{fig:cosmological_power_spectrum}
\end{figure}

@@

\begin{table*}[t]
\centering
\caption{Measured properties of Sculptor}
\label{tbl:Measured-properties-of-Sculptor}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012$˚ & M+18\\
$\delta$ & $-33.7186 \pm 0.00072$˚ & M+18\\
distance & $83.2 \pm 2$ kpc & Tran+22\\
$\mu_\alpha \cos \delta$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
RV & $111.2 \pm 0.2$ & arroyo-polonio+24\\
$\sigma_v$ & $9.7\pm0.2$ & arroyo-polonio+24\\
$R_h$ & $12.33 \pm 0.05$ arcmin & MV20*\\
$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & this work?\\
ell & $0.36 \pm 0.01$ & M+18\\
PA & $92\pm1$ & M+18\\
$M_V$ & $-10.82\pm0.14$ & M+18\\
\bottomrule
\end{tabular}
\end{table*}

Can I just cite \citet{arroyo-polonio+2024} for velocity information for
Sculptor? They correct for binary and do a more detailed discussion of
rotation.

@@

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. Composite Public
data/Giuseppe Donatiello. Updated image in 2022.
https://www.flickr.com/photos/133259498@N05/page8}\label{fig:umi_image}
\end{figure}

\begin{table*}[t]
\centering
\caption{Measured properties of Ursa Minor. References are: (1) Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020), (4) average of Pace et al. (2020) and Spencer et al. (2018).}
\label{tbl:Measured-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.00158$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ & 2\\
distance & $70.1 \pm 3.6$ kpc & \\
$\mu_\alpha \cos \delta$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
RV & $-244.9 \pm 0.3_{\rm stat} \pm 1.7_{\rm sys}$ km s$^{-1}$ & 4\\
$\sigma_v$ & $8.3 \pm 0.3_{\rm stat} \pm 0.4_{\rm sys}$ & 4\\
$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
ell & $0.55 \pm 0.01$ & 1\\
PA & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/idealized_break_radius.png}}
\caption[Break radius validation]{The break radius of the simulations is
set by}\label{fig:idealized_break_radius}
\end{figure}

@@

>>>>

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/power_spectrum.png}
\caption[Cosmological Power Spectrum]{The matter power spectrum under
different assumptions for dark matter. Dwarf galaxies occupy the middle
and low end of the blue region (10\^{}10 - 10\^{}8 solar masses),
enabling a unique window into properties of dark matter on small scales.
The smaller scales we can understand dark matter, the better we are able
to test different models of dark matter. figure 1 from
\citet{bechtol+2022}.}\label{fig:cosmological_power_spectrum}
\end{figure}

@@

\begin{table*}[t]
\centering
\caption{Observed properties of Sculptor. References are: 1. Muñoz et al. (2018), 2. Tran et al. (2022), 3. (MV2020b?), 4 McConnachie and Venn (2020).}
\label{tbl:Observed-properties-of-Sculptor-References-are:-1-Muñoz-et-al-2018-2-Tran-et-al-2022-3-MV2020b?-4-McConnachie-and-Venn-2020}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RRL) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $111.2 \pm 0.2\ {\rm km\,s^{-1}}$ & this work\\
$\sigma_v$ & $9.67\pm0.16\ {\rm km\,s^{-1}}$ & this work\\
$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & this work?\\
ellipticity & $0.36 \pm 0.01$ & 1\\
position angle & $92\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. UMi appears as a
diagonal/elliptical haze of stars from the top left to the bottom right.
Even as as classical dwarf, Ursa minor is fairly diffuse and does not
stand obviously out from the background. Composite Public data/Giuseppe
Donatiello. Updated image in 2022.
https://www.flickr.com/photos/133259498@N05/page8}\label{fig:umi_image}
\end{figure}

\begin{table*}[t]
\centering
\caption{Observed properties of Ursa Minor. References are: (1) Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020), (4) average of Pace et al. (2020) and Spencer et al. (2018).}
\label{tbl:Observed-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RRL) & 2\\
distance & $70.1 \pm 3.6$ kpc & -\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $-245.8 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & this work\\
$\sigma_v$ & $8.8 \pm 0.2$ & this work\\
$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/idealized_break_radius.png}}
\caption[Break radius validation]{The break radius of the simulations is
set by the time since pericentre.}\label{fig:idealized_break_radius}
\end{figure}

@@

\section{What is up with
probabilities}\label{what-is-up-with-probabilities}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250416112038183.png}
\caption{image-20250416112038183}\label{fig:test}
\end{figure}
